---
title: "R motivation: ggplot2"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

Do you have the most recent version of R and RStudio?
update all of your packages

```
version$version.string
RStudio.Version()$version
update.packages(ask = FALSE, checkBuilt = TRUE)

```

*R code example: Linear regression 

```{r}
fit = lm(mpg ~ wt, data = mtcars)
summary(fit)
```

* Baes Plot

```{r}
par(mar = c(4, 4, 1, .1)) ## Just for nice plot margins on this slide deck
plot(mtcars$wt, mtcars$mpg)
```

* ggplot

## More on ggplot2


If you don't have them already, install the `ggplot2` and `gapminder` packages via either:

*Console*: Enter `install.packages(c("ggplot2", "gapminder"), dependencies=T)`.

*RStudio*: Click the "Packages" tab in the bottom-right window pane. Then click "Install" and search for these two packages.

```{r}
library(ggplot2)
ggplot(data = mtcars, aes(x = wt, y = mpg)) + 
  geom_smooth(method = "lm", col = "red") + 
  geom_point()
```


##

Once the packages are installed, load them into your R session with the library() function.

```{r}
library(ggplot2)
library(gapminder) ## We're just using this package for the gapminder data
```
Notice too that you don't need quotes around the package names any more.

## The gapminder dataset

Because we're going to be plotting the gapminder dataset, it is helpful to know that it contains panel data on life expectancy, population size, and GDP per capita for 142 countries since the 1950s.

```{r}
gapminder
```

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
```

##

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, col = continent)) + facet_wrap(~continent)+
  geom_point(alpha = 0.3) ## "alpha" controls transparency. Takes a value between 0 and 1.
```
##

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, col = continent)) + scale_x_log10() +
  geom_point(alpha = 0.3)  + labs(title="y vs log(x)")
```

## intermediate plot object

At this point, instead of repeating the same ggplot2 call every time, it will prove convenient to define an intermediate plot object that we can re-use.

```{r}
p = ggplot(data = gapminder, 
           aes(x = gdpPercap, y = lifeExp)) + scale_x_log10() 
p
```

```{r}
p + 
  geom_point(alpha = 0.3)  +
  geom_smooth(method = "loess")
```

```{r}
p + 
  geom_point(aes(color=continent), alpha = 0.3)  +
  geom_smooth(method = "loess")
```
```{r}
ggplot(data = gapminder) + 
  geom_density(aes(x = gdpPercap, fill = continent), alpha=0.3)
```
```{r}
p2 =
  p + 
  geom_point(aes(col = continent), alpha = 0.3) +
  scale_x_log10(labels = scales::dollar) + ## Switch to logarithmic scale on x-axis. Use dollar units.
  labs(x = "Log (GDP per capita)", y = "Life Expectancy") + ## Better axis titles
  theme_minimal() ## Try a minimal (b&w) plot theme
p2
```


## animation
```{r}
library(gganimate)
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```

*Some useful reading

- Chapter 3 of R for Data Science by Hadley Wickham and Garett Grolemund.
- Data Visualization: A Practical Guide by Kieran Healy.
- Designing ggplots by Malcom Barrett.
